---
title: "Inspect JUMP-Target metadata and update gene-compound connections"
---

(From here <https://github.com/jump-cellpainting/JUMP-Target/issues/8#issuecomment-756712175>):

We use additional target annotations available from clue.io/repurposing to produce

1. An expanded set of target annotations: `output/JUMP-Target_compound_metadata_all_targets.csv`, 
along with two different versions of the same information that are easier to browse:  `output/JUMP-Target_compound_metadata_all_targets_per_compound.csv` and `output/JUMP-Target_compound_metadata_all_targets_per_target.csv`
2. An expanded set of connections between ORFs, CRISPRs, and compounds: `output/JUMP-Target_compounds_crispr_orf_connections.csv`, along with different 
version of the same information, stored as a gene x compound matrix: 
`output/JUMP-Target_gene_compound_matrix.csv`

# Setup

```{r message=FALSE}
library(magrittr)
library(tidyverse)
library(glue)
library(arrow)
```

# Read data

## Internal data

```{r}
drug_target_samples <-
  read_csv(
    "input/drug_target_samples.csv",
    col_types =  cols(
      pert_iname = col_character(),
      target = col_character(),
      broad_id = col_character()
    )
  ) %>%
  mutate(pert_id = str_sub(broad_id, 1, 13)) %>%
  rename(broad_sample = broad_id)
```

## External data

```{r}
commit <- "534520eb755ba374b2e1cd9ded2bc00050da93b6"
path <- glue("https://raw.githubusercontent.com/jump-cellpainting/JUMP-Target/{commit}")
```

Compounds

```{r}
compound_metadata <-
  read_tsv(
    glue("{path}/JUMP-Target-1_compound_metadata.tsv"),
    col_types = cols(
      broad_sample = col_character(),
      InChIKey = col_character(),
      pert_iname = col_character(),
      pubchem_cid = col_double(),
      target = col_character(),
      pert_type = col_character(),
      control_type = col_character(),
      smiles = col_character()
    )
  ) %>%
  mutate(pert_id = str_sub(broad_sample, 1, 13))
```

ORFs

```{r}
orf_metadata <-
  read_tsv(
    glue("{path}/JUMP-Target-1_orf_metadata.tsv"),
    col_types = cols(
      broad_sample = col_character(),
      gene = col_character(),
      pert_type = col_character(),
      control_type = col_character()
    )
  )
```

CRISPRs

```{r}
crispr_metadata <-
  read_tsv(
    glue("{path}/JUMP-Target-1_crispr_metadata.tsv"),
    col_types = cols(
      broad_sample = col_character(),
      gene = col_character(),
      pert_type = col_character(),
      control_type = col_character(),
      target_sequence = col_character(),
      negcon_control_type = col_character()
    )
  )
```

# Inspect and update metadata

## Drugs

```{r}
compound_pert_iname <- 
  compound_metadata %>% 
  select(pert_iname) %>%
  mutate(pert_iname = str_split(pert_iname, "\\|")) %>%
  unnest(cols = c(pert_iname)) %>%
  mutate(pert_iname = str_trim(pert_iname)) %>%
  distinct()
```

Number of unique `pert_iname`s

```{r}
compound_pert_iname %>%
  anti_join(data.frame(pert_iname = "DMSO")) %>%
  count() 
```

Create a complete mapping between `pert_id`s and `pert_iname`s for compounds, 
using `drug_target_samples` i.e. get all possible `pert_id` for a given
`pert_iname`.

```{r}
compound_pert_iname_pert_id <- 
  compound_pert_iname %>%
  distinct(pert_iname) %>%
  inner_join(drug_target_samples %>% 
               distinct(pert_id, pert_iname)) %>%
  distinct()
```

Number of pert_id-pert_iname pairs

```{r}
compound_pert_iname_pert_id %>%
  count()
```

`pert_id`s with multiple `pert_iname`s in the complete mapping

```{r}
compound_pert_iname_pert_id %>%
  add_count(pert_id) %>% 
  filter(n > 1)
```

`pert_iname`s with multiple `pert_id`s in the complete mapping

```{r rows.print=25}
compound_pert_iname_pert_id %>%
  add_count(pert_iname) %>% 
  filter(n > 1)
```

`pert_iname`s with multiple `pert_id`s in `JUMP-Target-1`

```{r rows.print=25}
compound_pert_iname_pert_id %>%
  inner_join(compound_metadata %>% distinct(pert_id)) %>%
  add_count(pert_iname) %>% 
  filter(n > 1)
```

As discussed here https://github.com/jump-cellpainting/JUMP-Target/issues/9#issuecomment-830667015
we will treat these as biological replicates.

So in all, we have these many unique compounds in the dataset, as far as training
models goes:

```{r}
compound_pert_iname %>%
  anti_join(data.frame(pert_iname = "DMSO")) %>%
  count() 
```

### Get full list of targets

List all genes present in the ORFs / CRISPRs in `JUMP-Target-1`

```{r}
jump_target_genes <- 
  bind_rows(
    crispr_metadata %>% distinct(gene),
    orf_metadata %>% distinct(gene),
  ) %>% 
  distinct(gene) %>%
  na.omit()
```


```{r}
jump_target_genes %>%
  count()
```

List compound `pert_id`s present in `JUMP-Target-1`

```{r}
jump_target_compound_pert_id <- 
  compound_metadata %>% 
  distinct(pert_id, pert_iname) %>%
  na.omit()
```


```{r}
jump_target_compound_pert_id %>%
  count()
```

Get `(pert_id, pert_iname, target)` map 

```{r}
compound_full_targets <-
  bind_rows(
    inner_join(
      compound_pert_iname_pert_id,
      drug_target_samples %>%
        distinct(pert_id, target),
      by = "pert_id"
    ),
    inner_join(
      compound_pert_iname_pert_id,
      drug_target_samples %>%
        distinct(pert_iname, target),
      by = "pert_iname"
    )
  ) %>%
  distinct(pert_id, pert_iname, target)
```


```{r}
compound_full_targets <-
  bind_rows(compound_full_targets,
            compound_metadata %>%
              distinct(pert_id, pert_iname, target)) %>%
  distinct(pert_id, pert_iname, target)
```

Drop targets not present in `JUMP-Target-1` ORFs/CRISPRs

```{r}
compound_full_targets <-
  compound_full_targets %>%
  inner_join(jump_target_genes, by = c("target" = "gene"))
```

Drop `pert_id`s not present in `JUMP-Target-1` Compounds

```{r rows.print=20}
compound_full_targets %>%
  anti_join(jump_target_compound_pert_id) %>%
  arrange(pert_iname)
```


```{r}
compound_full_targets <-
  compound_full_targets %>%
  distinct(pert_iname, target) %>%
  inner_join(jump_target_compound_pert_id, by = "pert_iname")
```


```{r}
compound_full_targets %>%
  count()
```
```{r}
compound_full_targets %>%
  distinct(pert_id, target) %>%
  count()
```


```{r}
compound_full_targets %>%
  distinct(pert_iname, target) %>%
  count()
```

```{r}
compound_full_targets %>%
  distinct(target) %>%
  count()
```
We investigate why this is not exactly 160, later.

```{r}
set.seed(42)
compound_full_targets %>% sample_n(20)
```


### Probe full list of targets

```{r}
compound_full_targets %>%
  distinct(pert_iname) %>%
  count()
```


```{r}
compound_full_targets %>%
  distinct(pert_id) %>%
  count()
```


```{r}
compound_full_targets %>%
  distinct(pert_iname, pert_id) %>%
  count()
```


```{r}
pert_id_pert_name_counts <- 
  compound_full_targets %>%
  distinct(pert_iname, pert_id) %>%
  add_count(pert_iname, name = "n_pert_iname")  %>%
  add_count(pert_id, name = "n_pert_id")
```

The compounds have the same name but different `pert_id`s. 
There are different reasons for why this can be the case, but we should treat 
them, essentially, as replicates when training predictive models.

```{r}
pert_id_pert_name_counts %>% 
  filter((n_pert_iname > 1) | (n_pert_id > 1)) %>%
  arrange(n_pert_iname, pert_iname, n_pert_id, pert_id)
```

### Save full list of targets

```{r}
compound_full_targets %>%
  select(pert_iname, pert_id, target) %>%
  arrange(pert_iname, target) %>%
  write_csv("output/JUMP-Target_compound_metadata_all_targets.csv")
```

```{r}
read_csv("output/JUMP-Target_compound_metadata_all_targets.csv", col_types = cols()) %>%
  arrange(pert_iname, target) %>%
  group_by(pert_iname) %>%
  summarize(target = paste0(target, collapse = "|")) %>%
  write_csv("output/JUMP-Target_compound_metadata_all_targets_per_compound.csv")
  
read_csv("output/JUMP-Target_compound_metadata_all_targets.csv", col_types = cols()) %>%
  arrange(target, pert_iname) %>%
  group_by(target) %>%
  summarize(pert_iname = paste0(pert_iname, collapse = "|")) %>%
  write_csv("output/JUMP-Target_compound_metadata_all_targets_per_target.csv")
```


## Connections among genes and compounds

Get the dictionaries

```{r}
compound_metadata_x <- 
  compound_full_targets %>% 
  distinct(pert_id, pert_iname, target) %>% 
  select(pert_id_compound = pert_id, pert_iname_compound = pert_iname, gene = target)

orf_metadata_x <- 
  orf_metadata %>% 
  distinct(broad_sample, gene) %>% 
  select(broad_sample_orf = broad_sample, gene)

crispr_metadata_x <- 
  crispr_metadata %>% 
  distinct(broad_sample, gene) %>% 
  select(broad_sample_crispr = broad_sample, gene) %>% 
  na.omit()
```

### Get connections

```{r}
connections <-
  compound_metadata_x %>% 
  inner_join(orf_metadata_x, by = "gene") %>% 
  inner_join(crispr_metadata_x, by = "gene") %>%
  select(gene, pert_iname_compound, pert_id_compound, broad_sample_orf, broad_sample_crispr)
```

### Probe connection

#### Number of genes

```{r}
connections %>%
  distinct(gene) %>%
  count
```

```{r}
compound_full_targets %>%
  distinct(target) %>%
  count()
```

What's the difference?

```{r}
extra_genes <- 
  compound_full_targets %>%
  distinct(target) %>% 
  anti_join(connections %>% distinct(gene), by = c("target" = "gene"))
```

These are gene(s) that were included as ORF negcons, and so there are
no corresponding CRISPR guides targeting them.

```{r}
compound_full_targets %>% inner_join(extra_genes)

crispr_metadata %>% inner_join(extra_genes, by = c("gene" = "target"))

orf_metadata %>% inner_join(extra_genes, by = c("gene" = "target"))
```

These genes below have CRISPRs / ORFs but not compounds targeting them
(these are all the so-called ORF negcons)

```{r}
jump_target_genes %>% 
  anti_join(connections %>% distinct(gene)) %>%
  left_join(orf_metadata)
```

#### Number of compound-ORF combinations

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(pert_iname_compound, pert_id_compound, broad_sample_orf) %>%
  count
```
NOT counting compounds with different `pert_id`s separately; this is the right 
way to do it for the application at hand i.e. training models that can predict 
gene-compound matches.

```{r}
connections %>%
  distinct(pert_iname_compound, broad_sample_orf) %>%
  count
```
#### Number of compound-CRISPR combinations

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(pert_iname_compound, pert_id_compound, broad_sample_crispr) %>%
  count
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
connections %>%
  distinct(pert_iname_compound, broad_sample_crispr) %>%
  count
```

#### Number of CRISPR-ORF combinations

```{r}
connections %>%
  distinct(broad_sample_orf, broad_sample_crispr) %>%
  count
```

#### Number of Gene-Compound combinations

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(pert_id_compound, pert_iname_compound, gene) %>%
  count
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
connections %>%
  distinct(pert_iname_compound, gene) %>%
  count
```

#### Number of  CRISPR-ORF-Compound combinations


```{r}
connections %>%
  count
```

```{r}
connections %>%
  distinct() %>%
  count
```

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(across(-all_of("pert_iname_compound"))) %>%
  count
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
connections %>%
  distinct(across(-all_of("pert_id_compound"))) %>%
  count
```

#### Tabulation of number of compounds targeting genes (ORF)

`n_orfs` have `n_compounds` targeting them

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(pert_id_compound, broad_sample_orf) %>%
  group_by(broad_sample_orf) %>%
  tally(name = "n_compounds") %>% 
  group_by(n_compounds) %>%
  tally(name = "n_orfs") %>%
  select(n_orfs, n_compounds) %>%
  arrange(desc(n_orfs))
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
connections %>%
  distinct(pert_iname_compound, broad_sample_orf) %>%
  group_by(broad_sample_orf) %>%
  tally(name = "n_compounds") %>% 
  group_by(n_compounds) %>%
  tally(name = "n_orfs") %>%
  select(n_orfs, n_compounds) %>%
  arrange(desc(n_orfs))
```


#### Tabulation of number of compounds targeting genes (CRISPR)

`n_crisprs` have `n_compounds` targeting them

Counting compounds with different `pert_id`s separately

```{r}
connections %>%
  distinct(pert_id_compound, broad_sample_crispr) %>%
  group_by(broad_sample_crispr) %>%
  tally(name = "n_compounds") %>% 
  group_by(n_compounds) %>%
  tally(name = "n_crisprs") %>%
  select(n_crisprs, n_compounds) %>%
  arrange(desc(n_crisprs))
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
connections %>%
  distinct(pert_iname_compound, broad_sample_crispr) %>%
  group_by(broad_sample_crispr) %>%
  tally(name = "n_compounds") %>% 
  group_by(n_compounds) %>%
  tally(name = "n_crisprs") %>%
  select(n_crisprs, n_compounds) %>%
  arrange(desc(n_crisprs))
```

### Save connections

```{r}
connections %>%
  arrange(gene, pert_iname_compound) %>%
  write_csv("output/JUMP-Target_compounds_crispr_orf_connections.csv")
```

```{r}
inner_join(
  connections %>%
    distinct(pert_iname_compound) %>%
    mutate(dummy = 1),
  connections %>%
    distinct(gene) %>%
    mutate(dummy = 1)
) %>%
  left_join(connections %>%
              distinct(gene, pert_iname_compound) %>%
              mutate(is_connected = 1)) %>%
  mutate(is_connected = replace_na(is_connected, 0)) %>%
  select(-dummy) %>%
  arrange(pert_iname_compound, gene) %>%
  pivot_wider(names_from = "gene", values_from = "is_connected") %>%
  write_tsv("output/JUMP-Target_gene_compound_matrix.txt")
```

