---
title: "Inspect metadata used in predictive models"
---

Here, we compare the connections used for predictive modeling
("current connections"), with the expanded list of connections
("updated connections") generated in `0.inspect-metadata/0.update-metadata.Rmd`  (`0.inspect-metadata/output/JUMP-Target_compounds_crispr_orf_connections.csv`)

See https://github.com/jump-cellpainting/genemod/issues/5#issuecomment-831236914

# Setup

```{r message=FALSE}
library(glue)
library(magrittr)
library(tidyverse)
source("utils.R")
```

# Read data

@danieljmichael said:

All of the possible splits we consider are in the `splits/` directory.

The naming convention I use is `{formulation}__{features}__{train|val|test}.csv`.

So as an example, `compound__crispr_orf__test.csv` contains the test set for the
leave out compound split with CRISPR+ORF features.

The binary_predictions directory contains the binary predictions and predicted
probabilities for each sample in the test set for each possible split.


```{r}
get_connections_from_splits <- function(formulation, features) {
  map_dfr(c("train", "test", "val"),
          function(split_set_i) {
            read_csv(
              glue(
                "../1.analysis/splits2/{formulation}__{features}__{split_set_i}.csv"
              ),
              col_types = cols()
            ) %>%
              process_metadata()
          }) %>%
    arrange(across(everything()))
}
```

## Get connections from the pair:crispr_orf configuration

```{r}
current_connections_raw <-
  get_connections_from_splits(formulation = "pair",
                              features = "crispr_orf"
    )
```

## Compare other configurations with the reference configuration

```{r}
formulation_features <-
  read_csv("input/formulation_features_transformer.csv",
           col_types = cols(.default = col_character()))
```


```{r}
inspect_formulation_features <- function(formulation_features_df) {

  current_connections_raw_i <-
    get_connections_from_splits(formulation = formulation_features_df$formulation,
                                features = formulation_features_df$features)

  common_cols <- intersect(names(current_connections_raw_i),
                           names(current_connections_raw))

  df_ref <- current_connections_raw %>% select(all_of(common_cols))
  df_i <- current_connections_raw_i %>% select(all_of(common_cols))

  equal_ref <- all(all.equal(df_ref, df_i) == TRUE)

  subset_ref <- df_i %>% anti_join(df_ref) %>% nrow() == 0

  data.frame(nrow = nrow(df_i), equal_ref = equal_ref, subset_ref = subset_ref)

}
```


```{r}
inpsect_formulation_features_df <-
  formulation_features %>%
  group_by(formulation, features) %>%
  summarise(inspect_formulation_features(cur_group()), .groups = "keep")
```


```{r rows.print=20}
inpsect_formulation_features_df %>%
  arrange(formulation, features)
```

# Compare current connections with updated connections

This is the updated list of gene-compound connections

```{r}
updated_connections <-
  read_csv("../0.inspect-metadata/output/JUMP-Target_compounds_crispr_orf_connections.csv",
           col_types = cols()) %>%
  select(pert_id_compound, broad_sample_orf, broad_sample_crispr, pert_iname_compound, gene)
```


```{r}
current_connections <-
  current_connections_raw %>%
  select(pert_id_compound, broad_sample_orf, broad_sample_crispr, y) %>%
  mutate(is_connected = y == 1) %>%
  select(-y)
```

Add `pert_iname_compound` and `gene` columns to `current_connections`

```{r}
current_connections <-
  current_connections %>%
  inner_join(updated_connections %>% distinct(pert_id_compound, pert_iname_compound)) %>%
  inner_join(updated_connections %>% distinct(broad_sample_orf, gene))
```

## Current connections

Counting compounds with different `pert_id`s separately

```{r}
current_connections %>%
  filter(is_connected) %>%
  count()
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
current_connections %>%
  filter(is_connected) %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  count()
```

```{r}
current_connections %>%
  filter(is_connected) %>%
  select(pert_iname_compound, gene) %>%
  distinct() %>%
  count()
```

```{r}
current_connections %>%
  filter(is_connected) %>%
  distinct(pert_iname_compound) %>%
  count()
```

```{r}
current_connections %>%
  filter(is_connected) %>%
  distinct(gene) %>%
  count()
```


## Updated connections

Counting compounds with different `pert_id`s separately

```{r}
updated_connections %>%
  count()
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
updated_connections %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  count()
```

NOT counting compounds with different `pert_id`s separately, and only focusing on
gene-compound connections (ignore the actual ORF and CRISPRs involved)

```{r}
updated_connections %>%
  select(pert_iname_compound, gene) %>%
  distinct() %>%
  count()
```

```{r}
updated_connections %>%
  distinct(pert_iname_compound) %>%
  count()
```


```{r}
updated_connections %>%
  distinct(gene) %>%
  count()
```

## Current ∩ Updated connections

Counting compounds with different `pert_id`s separately

```{r}
current_connections %>%
  filter(is_connected) %>%
  inner_join(updated_connections) %>%
  distinct() %>%
  count()
```

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
current_connections %>%
  filter(is_connected) %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  inner_join(updated_connections %>%
               select(-pert_id_compound) %>%
               distinct()) %>%
  distinct() %>%
  count()
```

## Extra connections

Are there any tuples marked as `is_connected` that are absent in the updated connections?

```{r}
current_connections %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  filter(is_connected) %>%
  anti_join(updated_connections %>%   select(-pert_id_compound) %>%
              distinct()) %>%
  distinct()
```

## Missing connections

Are there any tuples not marked as `is_connected` that are present in the updated connections?

NOT counting compounds with different `pert_id`s separately; this is the right way to do it.

```{r}
updated_connections %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  anti_join(
    current_connections %>%
      select(-pert_id_compound) %>%
      distinct() %>%
      filter(is_connected)
  )
```

Filter the list down to the gene-compound pairs (using `pert_iname` for compounds)

```{r}
updated_connections %>%
  select(-pert_id_compound) %>%
  distinct() %>%
  anti_join(
    current_connections %>%
      select(-pert_id_compound) %>%
      distinct() %>%
      filter(is_connected)
  ) %>%
  distinct(gene, pert_iname_compound) %>%
  arrange(gene)
```
