---
title: "Cosine baseline - supervised"
params:
  formulation: compound
  formulation_column: pert_iname_compound
---

# Setup

```{r message=FALSE}
library(glue)
library(magrittr)
library(tidyverse)
```

```{r}
stopifnot(params$formulation %in% c("gene", "compound", "pair"))
stopifnot(
  (
    params$formulation == "gene" &
      params$formulation_column == "gene"
  ) |
    (
      params$formulation == "compound" &
        params$formulation_column == "pert_iname_compound"
    ) |
    (
      params$formulation == "pair" &
        (
          params$formulation_column == "pert_iname_compound" |
            params$formulation_column == "gene"
        )
    )
)
```

trimmed makes
- gene -> gene
- pert_iname_compound -> compound

```{r}
formulation_column_trimmed <- str_remove(params$formulation_column, "pert_iname_")
```


```{r}
cat(yaml::as.yaml(params))
```

# Read profile cosine similarities

```{r}
sim_df <- arrow::read_parquet("output/standard_experiments_sim.parquet", show_col_types = FALSE)
```

Bug fix column name (otherwise it gets annoying later in this code)

```{r}
sim_df <-
  sim_df %>%
  rename_with(~str_replace(., "Metadata_genes", "Metadata_gene"), everything())
```


# Make predictions

Read splits for the `compound` or `gene` formulation; the feature set does not matter

```{r}
splits <-
  read_csv(glue("output/splits__{params$formulation}__crispr.csv.gz"), show_col_types = FALSE)
```

Save it as it's own splits file so that it can be read in later

Call this the `compound`  or `gene` feature set because it uses compound or gene similarity

```{r}
splits %>%
  write_csv(glue("output/splits__{params$formulation}__{formulation_column_trimmed}.csv.gz"))
```


```{r}
# get test split
splits_test <-
  splits %>%
  filter(split_set == "test") %>%
  select(-split_set)

# get train split
splits_train <-
  splits %>%
  filter(split_set == "train") %>%
  select(-split_set)
```

Augment the test split with connections from train:
For each gene-compound pair, get either

- all compounds connected to the gene in train (for leave-out-compound)
- all genes connected to the compound in train (for leave-out-gene)


```{r}
splits_test_aug <-
  splits_test %>%
  inner_join(
    splits_train,
    by = ifelse(
      params$formulation_column == "pert_iname_compound",
      "gene",
      "pert_iname_compound"
    ),
    suffix = c("", "_train")
  )

rm(list = c("splits_test", "splits_train", "splits"))
```

Get similarity between

- compounds (for leave-out-compound)
- genes (for leave-out-gene)

```{r}
col1 <- sym(glue("Metadata_{params$formulation_column}1"))
col2 <- sym(glue("Metadata_{params$formulation_column}2"))

sim_df_modality <-
  sim_df %>%
  filter(Metadata_modality1 == Metadata_modality2) %>%
  filter(Metadata_modality1 %in% ifelse(
    params$formulation_column == "pert_iname_compound",
    c("Compound"),
    c("ORF", "CRISPR")
  )) %>%
  group_by(!!col1,
           !!col2) %>%
  summarise(sim = mean(sim), .groups = "keep") %>%
  ungroup()
```

Rescale `sim` to stretch the range [-1,1] so that gene-gene and compound-compaund similarities are comparable.
Use rank to make it uniform

```{r}
sim_df_modality <-
  sim_df_modality %>%
  mutate(sim = 2 * rank(sim) / length(sim) - 1)
```

Augment splits_test_aug with compound similarities, and then compute weighted average of cosine similarity:

- TRUE connections (is_connected==TRUE) are given a weight of +1
- FALSE connections (is_connected==FALSE) are given a weight of -1

```{r}
join_cols <-
  c(setNames(
    nm = params$formulation_column,
    glue("Metadata_{params$formulation_column}1")
  ),
  setNames(
    nm = glue("{params$formulation_column}_train"),
    glue("Metadata_{params$formulation_column}2")
  ))

join_cols

splits_test_aug_modality <-
  splits_test_aug %>%
  inner_join(sim_df_modality, by = join_cols)
```

## Make predictions using averages of all neighbors

```{r}
splits_test_aug_modality_avg <-
  splits_test_aug_modality %>%
  group_by(
    is_connected,
    gene,
    pert_iname_compound,
    is_connected_train
  ) %>%
  summarize(sim = mean(sim), n = n(), .groups = "keep") %>%
  mutate(is_connected_train_sign = if_else(is_connected_train, 1, -1)) %>%
  group_by(is_connected, gene, pert_iname_compound) %>%
  # smallest value (=-2) is when average of TRUE connections is -1 and average of FALSE connections is 1
  # largest value (=2) is when average of TRUE connections is 1 and average of FALSE connections is -1
  # normalize accordingly i.e (2 + x) / 4
  summarize(y_prob = (2 + sum(is_connected_train_sign * sim)) / 4, .groups = "keep") %>%
  ungroup()
```

Estimate positive fraction (prior)

```{r}
y_pos_frac <- sum(splits_test_aug_modality_avg$is_connected)/nrow(splits_test_aug_modality_avg)
y_pos_thread <- quantile(splits_test_aug_modality_avg$y_prob, 1-y_pos_frac, names = FALSE)
```

Create prediction data frame

```{r}
prediction_df <-
  splits_test_aug_modality_avg %>%
  select(Metadata_gene = gene,
         Metadata_pert_iname_compound = pert_iname_compound,
         y_actual = is_connected,
         y_prob) %>%
  mutate(y_pred = y_prob > y_pos_thread)
```

## Save predictions

```{r}
prediction_df %>%
  write_csv(glue("../1.analysis/final_binary_predictions/cosineavg__{params$formulation}__{formulation_column_trimmed}.csv"))
```

## Make predictions using nearest-neighbor

```{r}
prediction_df <-
  splits_test_aug_modality %>%
  group_by(is_connected,
           gene,
           pert_iname_compound) %>%
  arrange(desc(sim)) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(y_prob = if_else(is_connected_train, 1, -1) * (sim + 1) / 4 + 0.5,
         y_pred = is_connected_train) %>%
  select(
    Metadata_gene = gene,
    Metadata_pert_iname_compound = pert_iname_compound,
    y_actual = is_connected,
    y_prob,
    y_pred
  )
```

Sanity check

```{r}
modality_col <- sym(paste0("Metadata_", params$formulation_column))

count_check_1 <-
  prediction_df %>%
  mutate(x = ifelse(y_pred, y_prob, 1 - y_prob)) %>%
  count(!!modality_col, x, name = "n_repeats") %>%
  count(n_repeats)

count_check_2 <-
  prediction_df %>%
  count(!!modality_col, name = "n_repeats") %>%
  count(n_repeats)

compare::compare(
  count_check_1,
  count_check_2,
  allowAll = T)
```

## Save predictions

```{r}
prediction_df %>%
  write_csv(glue("../1.analysis/final_binary_predictions/cosine1nn__{params$formulation}__{formulation_column_trimmed}.csv"))
```

## Make predictions using 5-nearest-neighbor

```{r}
prediction_df <-
  splits_test_aug_modality %>%
  group_by(is_connected,
           gene,
           pert_iname_compound) %>%
  arrange(desc(sim)) %>%
  slice_head(n = 5) %>%
  summarize(
    y_prob = any(is_connected_train) * 1.0,
    y_pred = any(is_connected_train),
    .groups = "keep"
  ) %>%
  ungroup() %>%
  select(
    Metadata_gene = gene,
    Metadata_pert_iname_compound = pert_iname_compound,
    y_actual = is_connected,
    y_prob,
    y_pred
  )
```

## Save predictions

```{r}
prediction_df %>%
  write_csv(glue("../1.analysis/final_binary_predictions/cosine5nn__{params$formulation}__{formulation_column_trimmed}.csv"))
```
