---
title: "Inspect split consistency"
---

# Setup

```{r message=FALSE}
library(glue)
library(magrittr)
library(tidyverse)
```

# Read data

```{r}
formulation_features <- 
  read_csv("input/formulation_features_transformer.csv", 
           col_types = cols(.default = col_character()))

```

```{r}
splits_df <-
  formulation_features %>%
  select(formulation, features) %>%
  pmap_dfr(function(formulation, features) {
    splits <-
      read_csv(glue("output/splits__{formulation}__{features}.csv.gz"),
               col_types = cols())  %>%
      arrange(across(everything())) %>%
      mutate(formulation = formulation,
             features = features) %>%
      distinct()
  })
```


```{r}
splits_hash_df <-
  formulation_features %>%
  select(formulation, features) %>%
  pmap_dfr(function(formulation, features) {
    splits <-
      read_csv(glue("output/splits__{formulation}__{features}.csv.gz"),
               col_types = cols())  %>%
      arrange(across(everything()))
    
    hash <- str_sub(digest::digest(splits), 1, 8)
    
    tibble(formulation,
           features,
           hash)
  })
```

# Inspect

```{r rows.print=20}
splits_df %>%
  group_by(formulation, features, split_set) %>%
  count() %>%
  arrange(formulation, features, split_set) %>%
  pivot_wider(names_from = "split_set", values_from = n)
```

```{r}
splits_df %>%
  group_by(formulation, features, split_set) %>%
  count() %>%
  ggplot(aes(features, n, fill = split_set)) + geom_col() + facet_wrap(~formulation)
```

```{r rows.print=20}
splits_df %>%
  group_by(formulation, features, split_set) %>%
  count() %>%
  arrange(formulation, features, split_set) %>%
  pivot_wider(names_from = "features", values_from = n)
```

```{r}
splits_df %>%
  group_by(formulation, features, split_set) %>%
  count() %>%
  ggplot(aes(split_set, n, fill = features)) + geom_col(position = "dodge") + facet_wrap(~formulation)
```

```{r rows.print=20}
splits_hash_df %>% arrange(formulation, features, hash)
```


```{r rows.print=20}
splits_hash_df %>% arrange(hash, formulation, features)
```

Check this:

> In the leave-out-compound formulation, we know that the train and test sets have non-overlapping sets of compounds. So a test gene can be seen with a different compound during training which may or may not be a "sister compound" of the test compound. Similarly, in leave-out-compound-together, a test gene can be seen with a different compound during training, but we can guarantee the training compound is NOT a sister compound of the test compound. As a reminder, sister compounds refer to compounds that are distinct but share a gene target.

```{r rows.print=20}
# get splits for a formulation-features pair, drop validation set
splits_i <-
  splits_df %>%
  ungroup %>%
  inner_join(
    data.frame(
      classifier = "transformer",
      formulation = "compound_together",
      features = "orf"
    )
  ) %>%
  filter(is_connected) %>%
  filter(split_set != "val")

# get list of genes that are in a true connection in both train and test
gene_list <-
  splits_i %>%
  distinct(gene, split_set) %>%
  count(gene) %>%
  filter(n > 1) %>%
  select(-n)

# for each gene, list the true connections in test set and in train set
splits_i %>%
  select(gene, split_set, pert_iname_compound) %>%
  inner_join(gene_list, by = "gene") %>%
  arrange(gene, split_set, pert_iname_compound) %>%
  group_by(gene, split_set) %>%
  summarize(pert_iname_compounds = str_c(pert_iname_compound, collapse = ","))
```

