---
title: "Inspect granular"
params:
  percentile_threshold: 0.02
---

# Setup

```{r message=FALSE}
library(glue)
library(magrittr)
library(tidyverse)
set.seed(42)
```


```{r}
palette_manual_dark_2 <- c("#1B9E77", "#D95F02", "#E6AB02", "#7570B3")
```

# Read

```{r message=FALSE}
prediction_metrics_at_k_abs_compound <-
  read_csv("output/prediction_metrics_at_k_abs_compound.csv", show_col_types = FALSE)

prediction_metrics_at_k_abs_gene <-
  read_csv("output/prediction_metrics_at_k_abs_gene.csv", show_col_types = FALSE)

prediction_df <-
  read_csv("output/prediction_df.csv.gz", show_col_types = FALSE)
```

Report positive prevalance and use that to set `params$percentile_threshold`

(Set it to 2x of `pp_median` i.e. twice the number of hits)

```{r}
prediction_df %>%
  filter(classifier %in% c("transformer", "transformerx")) %>%
  mutate(y_actual = y_actual == 1) %>%
  group_by(classifier, formulation, features, y_actual) %>%
  count() %>%
  pivot_wider(names_from = "y_actual",
              names_prefix = "y_actual_",
              values_from = "n") %>%
  mutate(pp = y_actual_TRUE / (y_actual_TRUE + y_actual_FALSE)) %>%
  ungroup() %>%
  summarise(across(c("pp"), c(
    median = median, min = min, max = max
  )))
```


```{r}
prediction_metrics_at_k_abs_compound %>% slice_sample(n = 10)
prediction_metrics_at_k_abs_gene %>% slice_sample(n = 10)
prediction_df %>% slice_sample(n = 10)
```

# Process

## Predictions

Calculate these:

- `gene_rank`: the rank of the gene, in the ranked list of similarities of all genes to the compound
- `compound_rank`: the rank of the compound, in the ranked list of similarities of all compounds to the gene
- `gene_pct`: `gene_rank` as a percentile
- `compound_pct`: `compound_rank` as a percentile
- `gene_compound_global_rank`: the percentile of `y_prob` for that classifier-feature-formulation
- `gene_compound_local_pct`: (`gene_pct` +  `compound_pct`)/2

```{r}
prediction_df <-
  prediction_df %>%
  arrange(classifier, formulation, features, pert_iname_compound, desc(y_prob)) %>%
  group_by(classifier, formulation, features, pert_iname_compound) %>%
  mutate(gene_rank = row_number(-y_prob)) %>%
  mutate(gene_pct = gene_rank / n()) %>%
  ungroup() %>%
  arrange(classifier, formulation, features, gene, desc(y_prob)) %>%
  group_by(classifier, formulation, features, gene) %>%
  mutate(compound_rank = row_number(-y_prob)) %>%
  mutate(compound_pct = compound_rank / n()) %>%
  ungroup() %>%
  group_by(classifier, formulation, features) %>%
  mutate(gene_compound_global_rank = row_number(-y_prob)) %>%
  mutate(gene_compound_global_pct = gene_compound_global_rank / n()) %>%
  ungroup() %>%
  mutate(gene_compound_local_pct = (gene_pct + compound_pct) / 2)
```

```{r}
prediction_df %>%
  write_csv("output/prediction_df_augmented.csv.gz")
```

## Splits

Get data splits for the configuration

```{r message=FALSE}
splits <-
  prediction_df %>%
  distinct(classifier, formulation, features) %>%
  arrange(classifier, formulation, features) %>%
  pmap_dfr(function(classifier, formulation, features) {
    config <- tibble(classifier, formulation, features)

    config_trim <-
      config %>%
      mutate(features = str_remove_all(features, "__max_consensus")) %>%
      mutate(features = str_remove_all(features, "__consensus"))

    read_csv(with(
      config_trim,
      glue("output/splits__{formulation}__{features}.csv.gz")
    ),
    show_col_types = FALSE) %>%
      #filter(is_connected) %>%
      inner_join(config, by = character())

  }) %>%
  arrange(classifier, formulation, features, gene, pert_iname_compound)
```


```{r}
splits %>%
  write_csv("output/splits_collated.csv.gz")
```


```{r message=FALSE}
splits %>% slice_sample(n = 10)
```


```{r}
splits %>%
  filter(classifier %in% c("transformer", "transformerx") &
           formulation == "pair" &
           features == "orf") %>%
  distinct(split_set, gene) %>%
  count(split_set)

splits %>%
  filter(classifier %in% c("transformer", "transformerx") &
           formulation == "pair" &
           features == "orf") %>%
  distinct(split_set, pert_iname_compound) %>%
  count(split_set)
```

Get counts

`n_connected_compounds_train`: Number of compounds to which the gene (from the test set) is connected in the training set

```{r}
gene_train_count <-
  splits %>%
  filter(split_set == "test") %>%
  distinct(classifier, formulation, features, gene) %>%
  left_join(
    splits %>%
      filter(split_set == "train") %>%
      mutate(
        is_connected = as.factor(is_connected),
        gene = as.factor(gene)
      ) %>%
      group_by(
        classifier,
        formulation,
        features,
        gene,
        is_connected,
        .drop = FALSE
      ) %>%
      summarise(n_connected_compounds_train = n(),
                connected_compounds_train =
                         paste0(pert_iname_compound, collapse = "|"),
                .groups = "keep") %>%
      mutate(is_connected = as.logical(is_connected)),
    by = c("classifier", "formulation", "features", "gene")
  ) %>%
  mutate(n_connected_compounds_train = replace_na(n_connected_compounds_train, 0),
         connected_compounds_train =
           case_when(
             is_connected ~ connected_compounds_train,
             TRUE ~ ""
           ))

gene_train_count %>% slice_sample(n = 5)
```

All genes should have two sets of counts (for `is_connected` == `TRUE` and `is_connected` == `FALSE`)

```{r}
gene_train_count %>% count(classifier, formulation, features, gene) %>% filter(n != 2)
```


```{r}
gene_train_count %>%
  count(classifier, formulation, features, n_connected_compounds_train, is_connected, name = "n_genes") %>%
  select(classifier, formulation, features, n_genes, n_connected_compounds_train, is_connected) %>%
  filter(classifier == "transformer" & is_connected) %>%
  ggplot(aes(as.factor(n_connected_compounds_train), n_genes, fill = features)) +
  geom_col(position = "dodge") +
  facet_wrap(~formulation) +
  xlab("Number of compounds to which the gene (from the test set)\nis connected in the training set (TRUE connections)") +
  ylab("Number of genes")
```

```{r}
gene_train_count %>%
  filter(classifier == "transformer" & !is_connected) %>%
  ggplot(aes(n_connected_compounds_train, fill = features)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  facet_wrap(~formulation, scales = "free") +
  xlab("Number of compounds to which the gene (from the test set)\nis connected in the training set (FALSE connections)") +
  ylab("Number of genes")
```

`n_connected_genes_train`: Number of genes to which the compound (from the test set) is connected in the training set

```{r}
compound_train_count <-
  splits %>%
  filter(split_set == "test") %>%
  distinct(classifier, formulation, features, pert_iname_compound) %>%
  left_join(
    splits %>%
      filter(split_set == "train")  %>%
      mutate(
        is_connected = as.factor(is_connected),
        pert_iname_compound = as.factor(pert_iname_compound)
      ) %>%
      group_by(
        classifier,
        formulation,
        features,
        pert_iname_compound,
        is_connected,
        .drop = FALSE
      ) %>%
      summarize(n_connected_genes_train = n(),
                connected_genes_train =
                         paste0(gene, collapse = "|"),
                .groups = "keep") %>%
      mutate(is_connected = as.logical(is_connected))
  ) %>%
  mutate(n_connected_genes_train = replace_na(n_connected_genes_train, 0),
         connected_genes_train =
           case_when(
             is_connected ~ connected_genes_train,
             TRUE ~ ""
           ))

compound_train_count %>% slice_sample(n = 5)
```


```{r}
compound_train_count %>%
  count(classifier, formulation, features, n_connected_genes_train, is_connected, name = "n_compounds") %>%
  select(classifier, formulation, features, n_compounds, n_connected_genes_train, is_connected) %>%
  filter(classifier == "transformer" & is_connected) %>%
  ggplot(aes(as.factor(n_connected_genes_train), n_compounds, fill = features)) +
  geom_col(position = "dodge") +
  facet_wrap(~formulation) +
  xlab("Number of genes to which the compound (from the test set)\nis connected in the training set (TRUE connections)") +
  ylab("Number of compounds")
```


```{r}
compound_train_count %>%
  filter(classifier == "transformer" & !is_connected) %>%
  ggplot(aes(n_connected_genes_train, fill = features)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  facet_wrap(~formulation, scales = "free") +
  xlab("Number of genes to which the compound (from the test set)\nis connected in the training set (FALSE connections)") +
  ylab("Number of compounds")
```

## Bias

For *positive* connections in the test set, report

- `n_connected_compounds_train`: Number of compounds to which the gene connected, in the training set
- `n_connected_genes_train`: Number of genes to which the compound is connected, in the training set
- `gene_rank`: the rank of the gene, in the ranked list of similarities of all genes to the compound, in the training set
- `compound_rank`: the rank of the compound, in the ranked list of similarities of all compounds to the gene, in the training set
- `gene_pct`: `gene_rank` as a percentile
- `compound_pct`: `gene_rank` as a percentile
- `gene_compound_local_pct`: (`gene_pct` +  `compound_pct`)/2
- `any_connected_train`: (`n_connected_genes_train` + `n_connected_compounds_train`) > 0
- `which_connected_train`: is the gene or the compound, or both, present in the training set?

```{r}
training_bias <-
  prediction_df %>%
  filter(y_actual == 1) %>%
  select(-y_actual) %>%
  inner_join(
    compound_train_count %>% filter(is_connected) %>% select(-is_connected),
    by = c(
      "classifier",
      "formulation",
      "features",
      "pert_iname_compound"
    )
  ) %>%
  inner_join(
    gene_train_count %>% filter(is_connected) %>% select(-is_connected),
    by = c(
      "classifier",
      "formulation",
      "features",
      "gene")
  ) %>%
  mutate(any_connected_train =
           (n_connected_genes_train + n_connected_compounds_train) > 0) %>%
  mutate(
    which_connected_train =
      case_when(
        n_connected_genes_train == 0 &
          n_connected_compounds_train == 0 ~ "neither",
        n_connected_genes_train == 0 &
          n_connected_compounds_train > 0 ~ "compound",
        n_connected_genes_train > 0 &
          n_connected_compounds_train == 0 ~ "gene",
        n_connected_genes_train > 0 &
          n_connected_compounds_train > 0 ~ "both"
      )
  ) %>%
  arrange(formulation, features, classifier, pert_iname_compound, gene)
```

These values should be the same

```{r}
prediction_df %>%
  filter(y_actual == 1) %>%
  count()

training_bias %>%
  count()
```

The number of connections should be the same

```{r}
prediction_df %>%
  filter(features != "compound") %>%
  count(classifier, formulation, features, y_actual) %>%
  pivot_wider(names_from = "classifier", values_from = "n") %>%
  filter(cosine != transformer)
```


```{r}
training_bias %>% slice_sample(n = 5)
```

```{r}
training_bias %>%
  write_csv("output/training_bias.csv")
```

# Inspect

## Relationships among scores

```{r}
training_bias <-
  training_bias %>%
  filter(classifier %in% c("transformer", "transformerx", "cosine", "cosine1nn", "cosine5nn", "cosineavg")) %>%
  mutate(features = str_replace(features, "__max_consensus", "_max")) %>%
  mutate(features = str_replace(features, "__consensus", "_avg"))
```

Inspect relationship between `gene_compound_local_pct` and `y_prob`

```{r}
training_bias %>%
  mutate(features = str_replace(features, "__", " ")) %>%
  group_by(classifier, formulation, features) %>%
  mutate(y_prob_rank = rank(y_prob),
         gene_compound_local_pct_rank = rank(gene_compound_local_pct)) %>%
  ggplot(aes(y_prob_rank, gene_compound_local_pct_rank, color = classifier)) +
  geom_point(size = .5) +
  facet_grid(features ~ formulation, labeller =  label_wrap_gen(multi_line = FALSE))
```

Inspect relationship between `gene_compound_local_pct` and `gene_compound_global_pct`

```{r}
training_bias %>%
  mutate(features = str_replace(features, "__", " ")) %>%
  group_by(classifier, formulation, features) %>%
  mutate(gene_compound_global_pct_rank = rank(gene_compound_global_pct),
         gene_compound_local_pct_rank = rank(gene_compound_local_pct)) %>%
  ggplot(aes(gene_compound_global_pct_rank, gene_compound_local_pct_rank, color = classifier)) +
  geom_point(size = .5) +
  facet_grid(features ~ formulation, labeller =  label_wrap_gen(multi_line = FALSE))
```

Inspect relationship between `y_prob` and `gene_compound_global_pct`

```{r}
training_bias %>%
  mutate(features = str_replace(features, "__", " ")) %>%
  group_by(classifier, formulation, features) %>%
  mutate(gene_compound_global_pct_rank = rank(gene_compound_global_pct),
         y_prob_rank = rank(y_prob)) %>%
  ggplot(aes(gene_compound_global_pct_rank, y_prob_rank, color = classifier)) +
  geom_point(size = .5) +
  facet_grid(features ~ formulation, labeller =  label_wrap_gen(multi_line = FALSE))
```
```{r}
training_bias %>%
  mutate(features = str_replace(features, "__", " ")) %>%
  ggplot(aes(gene_compound_global_pct, y_prob, color = classifier)) +
  geom_point(size = .5) +
  facet_grid(features ~ formulation, labeller =  label_wrap_gen(multi_line = FALSE))
```


```{r}
training_bias %>%
  group_by(classifier, formulation, features) %>%
  count()
```

## Counts of connection categories

```{r}
counts <-
  training_bias %>%
  filter(classifier == "transformer" & features == "orf") %>%
  count(formulation, which_connected_train) %>%
  arrange(formulation, desc(which_connected_train)) %>%
  group_by(formulation) %>%
  mutate(label_y = cumsum(n))

p <-
  counts %>%
  ggplot(aes(formulation, n, fill = which_connected_train)) +
  geom_col() +
  geom_text(aes(label = n, y = label_y), vjust = 1.5) +
  scale_fill_manual(values = palette_manual_dark_2) +
  ylab("Number of true test set connections") +
  ggtitle("Training bias: counts of true test set connections",
          subtitle = "(based on ORF features, so only one connection for each gene-compound pair)")

ggsave(
  glue("output/training_bias_counts.png"),
  p,
  width = 7,
  height = 5
)

print(p)
```

## Distribution of scores split by connection categories

### Continuous

```{r}
plot_training_global_a <- function(classifier_, prediction_type_) {

  if (prediction_type_ == "global_percentile") {
    prediction_col <- sym("gene_compound_global_pct")
    ylab_text <- "Connection percentile\n(test)"
  } else if (prediction_type_ == "local_percentile") {
    prediction_col <- sym("gene_compound_local_pct")
    ylab_text <- "Average percentile of the connection\namong all gene-compound pairs\n(test)"
  } else if (prediction_type_ == "prob") {
    prediction_col <- sym("y_prob")
    ylab_text <- "Predicted probability of \nthe connection being true\n(test)"
  } else {
    stop(glue("prediction_type {prediction_type_} unknown"))
  }

  p <-
    training_bias %>%
    filter(classifier == classifier_) %>%
    ggplot(aes(features, !!prediction_col, fill = which_connected_train)) +
    geom_dotplot(
      binaxis = "y",
      stackdir = "center",
      dotsize = 1.5,
      stroke = 0,
      position = "dodge",
      binpositions = "all",
      stackgroups = TRUE,
      binwidth = .01
    ) +
    labs(caption = "Each point is a true gene-compound connection (test)") +
    ylab(ylab_text) +
    theme(
      plot.caption = element_text(face = "italic"),
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1
      ),
      legend.position = "bottom"
    ) +
    scale_fill_manual(values = palette_manual_dark_2) +
    facet_wrap( ~ formulation) +
    guides(fill = guide_legend(title = "Gene or compound present in train\nin a true gene-compound connection")) +
    ggtitle(glue("Training bias: {classifier_} : {prediction_type_}"),
            subtitle = "Predictions are significantly worse if gene or compound was not seen in training\nin a true gene-compound connection (in supervised scenario)")

  ggsave(
    glue("output/training_bias_a_{prediction_type_}_{classifier_}.png"),
    p,
    width = 8,
    height = 5
  )

  p
}
```


```{r}
plot_training_global_a("transformer", prediction_type_ = "global_percentile")
plot_training_global_a("transformer", prediction_type_ = "local_percentile")
plot_training_global_a("transformer", prediction_type_ = "prob")
plot_training_global_a("transformerx", prediction_type_ = "global_percentile")
plot_training_global_a("transformerx", prediction_type_ = "local_percentile")
plot_training_global_a("transformerx", prediction_type_ = "prob")
plot_training_global_a("cosine", prediction_type_ = "global_percentile")
plot_training_global_a("cosine", prediction_type_ = "local_percentile")
plot_training_global_a("cosine1nn", prediction_type_ = "global_percentile")
plot_training_global_a("cosine1nn", prediction_type_ = "local_percentile")
plot_training_global_a("cosine5nn", prediction_type_ = "global_percentile")
plot_training_global_a("cosine5nn", prediction_type_ = "local_percentile")
plot_training_global_a("cosineavg", prediction_type_ = "global_percentile")
plot_training_global_a("cosineavg", prediction_type_ = "local_percentile")
```

### Thresholded

```{r}
training_bias_contingency <-
  training_bias %>%
  mutate(is_top_global_pct = as.factor(gene_compound_global_pct < params$percentile_threshold)) %>%
  mutate(is_top_local_pct = as.factor(gene_compound_local_pct < params$percentile_threshold)) %>%
  mutate(any_connected_train = as.factor(any_connected_train))

training_bias_contingency_global <-
  training_bias_contingency %>%
  group_by(classifier, formulation, features, .drop = FALSE) %>%
  count(any_connected_train, is_top_global_pct, .drop = FALSE) %>%
  rename(is_top_pct = is_top_global_pct) %>%
  mutate(prediction_type = "global_percentile")

training_bias_contingency_local <-
  training_bias_contingency %>%
  group_by(classifier, formulation, features, .drop = FALSE) %>%
  count(any_connected_train, is_top_local_pct, .drop = FALSE) %>%
  rename(is_top_pct = is_top_local_pct) %>%
  mutate(prediction_type = "local_percentile")

training_bias_contingency <-
  bind_rows(
    training_bias_contingency_global,
    training_bias_contingency_local
  )
```


```{r}
plot_training_global_b <- function(classifier_, prediction_type_) {
  p <-
    training_bias_contingency %>%
    filter(classifier == classifier_ & prediction_type == prediction_type_) %>%
    group_by(classifier, formulation, features) %>%
    mutate(frac = n / sum(n)) %>%
    ungroup() %>%
    ggplot(aes(features,
               frac,
               fill = is_top_pct,
               alpha = any_connected_train)) +
    geom_col() +
    facet_wrap( ~ formulation) +
    scale_fill_manual(values = c("#ef8a62", "#67a9cf")) +
    scale_alpha_manual(values = c(0.7, 1)) +
    guides(
      alpha = guide_legend(title = "Gene or compound\npresent in train\nin a true gene-compound\nconnection"),
      fill =
        guide_legend(title =
                       with(
                         params,
                         glue("Connection is\nin top {percentile_threshold}")
                       ))
    ) +
    ylab("Fraction of true connections") +
    theme(
      plot.caption = element_text(face = "italic"),
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1
      )
    ) +
    ggtitle(glue("Training bias: {classifier_}:{prediction_type_}"),
            subtitle = "Predictions are significantly worse if gene or compound was not seen in training\nin a true gene-compound connection (in supervised scenario)")

  ggsave(
    glue("output/training_bias_b_{prediction_type_}_{classifier_}.png"),
    p,
    width = 7,
    height = 6
  )

  p
}
```


```{r}
plot_training_global_b("transformer", prediction_type_ = "global_percentile")
plot_training_global_b("transformer", prediction_type_ = "local_percentile")
plot_training_global_b("transformerx", prediction_type_ = "global_percentile")
plot_training_global_b("transformerx", prediction_type_ = "local_percentile")
plot_training_global_b("cosine", prediction_type_ = "global_percentile")
plot_training_global_b("cosine", prediction_type_ = "local_percentile")
plot_training_global_b("cosine1nn", prediction_type_ = "global_percentile")
plot_training_global_b("cosine1nn", prediction_type_ = "local_percentile")
plot_training_global_b("cosine5nn", prediction_type_ = "global_percentile")
plot_training_global_b("cosine5nn", prediction_type_ = "local_percentile")
plot_training_global_b("cosineavg", prediction_type_ = "global_percentile")
plot_training_global_b("cosineavg", prediction_type_ = "local_percentile")
```


```{r}
plot_training_global_c <- function(classifier_, prediction_type_) {

  if (prediction_type_ == "global_percentile") {
    prediction_col <- sym("gene_compound_global_pct")
  } else if (prediction_type_ == "local_percentile") {
    prediction_col <- sym("gene_compound_local_pct")
  } else {
    stop(glue("prediction_type {prediction_type_} unknown"))
  }

  p <-
    training_bias %>%
    filter(classifier == classifier_) %>%
    group_nest(classifier, formulation, features) %>%
    mutate(neglogp = map(data,
                         function(df) {
                           -log10(wilcox.test(!!prediction_col ~ any_connected_train, data = df)$p.value)
                         })) %>%
    unnest(neglogp) %>%
    select(-data) %>%
    ggplot(aes(features, neglogp)) +
    geom_col() +
    geom_label(aes(label = round(neglogp, 1))) +
    ylab("-log10 p-value") +
    facet_wrap( ~ formulation) +
    theme(
      plot.caption = element_text(face = "italic"),
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1
      )
    ) +
    labs(caption = "Mann-Whitney test comparing predictions when gene or compound is seen vs. not-seen in training") +
    ggtitle(glue("Training bias: {classifier_}:{prediction_type_}"),
            subtitle = "Predictions are significantly worse if gene or compound was not seen in training\nin a true gene-compound connection (in supervised scenario)")

  ggsave(
    glue("output/training_bias_c_{prediction_type_}_{classifier_}.png"),
    p,
    width = 6,
    height = 6
  )

  p
}
```

### P-values

```{r}
plot_training_global_c("transformer", prediction_type_ = "global_percentile")
plot_training_global_c("transformer", prediction_type_ = "local_percentile")
plot_training_global_c("transformerx", prediction_type_ = "global_percentile")
plot_training_global_c("transformerx", prediction_type_ = "local_percentile")
plot_training_global_c("cosine", prediction_type_ = "global_percentile")
plot_training_global_c("cosine", prediction_type_ = "local_percentile")
plot_training_global_c("cosine1nn", prediction_type_ = "global_percentile")
plot_training_global_c("cosine1nn", prediction_type_ = "local_percentile")
plot_training_global_c("cosine5nn", prediction_type_ = "global_percentile")
plot_training_global_c("cosine5nn", prediction_type_ = "local_percentile")
plot_training_global_c("cosineavg", prediction_type_ = "global_percentile")
plot_training_global_c("cosineavg", prediction_type_ = "local_percentile")
```


