---
title: "Inspect split consistency (again)"
---

# Setup

```{r message=FALSE}
library(glue)
library(magrittr)
library(tidyverse)
```

# Why we no longer use leave-out-compound-together

## Read splits

```{r}
formulation_features <-
  read_csv("input/formulation_features_transformer.csv",
           col_types = cols(.default = col_character()))

```

```{r}
splits_df <-
  formulation_features %>%
  select(formulation, features) %>%
  pmap_dfr(function(formulation, features) {
    splits <-
      read_csv(glue("output/splits__{formulation}__{features}.csv.gz"),
               col_types = cols())  %>%
      arrange(across(everything())) %>%
      mutate(formulation = formulation,
             features = features) %>%
      distinct()
  })
```

## Check

Check if this holds up:

> In the leave-out-compound formulation, we know that the train and test sets have non-overlapping sets of compounds. So a test gene can be seen with a different compound during training which may or may not be a "sister compound" of the test compound. Similarly, in leave-out-compound-together, a test gene can be seen with a different compound during training, but we can guarantee the training compound is NOT a sister compound of the test compound. As a reminder, sister compounds refer to compounds that are distinct but share a gene target.

Get splits for a formulation-features pair, drop validation set (we only care about traina and test)

```{r}
splits_x <-
  splits_df %>%
  ungroup %>%
  inner_join(
    data.frame(
      classifier = "transformer",
      formulation = "compound_together",
      features = "orf"
    )
  ) %>%
  filter(is_connected) %>%
  filter(split_set != "val")
```

Get list of genes that are in a true connection in both train and test

```{r}
gene_list <-
  splits_x %>%
  distinct(gene, split_set) %>%
  count(gene) %>%
  filter(n > 1) %>%
  select(-n)
```

For each gene, list the true connections in test set and in train set

```{r}
splits_x %>%
  select(gene, split_set, pert_iname_compound) %>%
  inner_join(gene_list, by = "gene") %>%
  arrange(gene, split_set, pert_iname_compound) %>%
  group_by(gene, split_set) %>%
  summarize(pert_iname_compounds = str_c(pert_iname_compound, collapse = ","))
```

Regarding compound-together: I think I know what's going on. It looks like the definition of "sister compound" uses the target annotations that were available in the normalized_feature_select_negcon.csv file that [we](https://github.com/jump-cellpainting/genemod/blob/e0e2047752d2152c94893c02ccd342d2b4a7139f/1.analysis/code/utils_stanford.py#L112) load (which in turn gets its annotations from [here](https://github.com/jump-cellpainting/pilot-cpjump1-analysis/blob/e9e5b7d3ea0d5b8e9a162598fa838b036976e8e0/1.cpjump1-analysis/input/JUMP-Target_compound_metadata.tsv)). This file contains only a single target per compound vs multiple targets per compound that we [use](https://github.com/jump-cellpainting/genemod/blob/e0e2047752d2152c94893c02ccd342d2b4a7139f/1.analysis/code/utils_stanford.py#L41-L46) in the rest of our analysis.

The single target picked is [arbitrary](https://github.com/jump-cellpainting/pilot-cpjump1-analysis/issues/4#issuecomment-1024138895), so we will compound-together from our analysis completely, because it's essentially not very different from leave-out-compound (just that there are fewer sisters, but not no sisters, across train and test). Thankfully, leave-out-gene provides a very similar configuration (I think), so we don't lose out much.

# Check leave-out-pair

## Read original gene-compound matrix

```{r}
x0 <-
  read_tsv("../0.inspect-metadata/output/JUMP-Target_gene_compound_matrix.txt") %>%
  pivot_longer(-pert_iname_compound,
               names_to = "gene",
               values_to = "is_connected") %>%
  mutate(is_connected = is_connected == 1) %>%
  mutate(pert_iname_compound = as.factor(pert_iname_compound),
         gene = as.factor(gene))
```


```{r}
x0 %>%
  count(is_connected)
```
## Specify counts in the specific split we use

```{r}
training_bias_counts_genemod <-
  data_frame(both = 42, compound = 32, gene =  5, neither = 23)
```


```{r}
splits_genemod <-
  splits_df %>%
  ungroup %>%
  inner_join(
    data.frame(
      classifier = "transformer",
      formulation = "pair",
      features = "orf"
    )
  ) %>%
  select(split_set, is_connected, gene, pert_iname_compound) %>%
  arrange(split_set, is_connected, gene, pert_iname_compound) %>%
    mutate(across(c(gene, pert_iname_compound), as.factor))

```

```{r}
create_splits <- function(x) {
  x %>%
    mutate(s = runif(nrow(.))) %>%
    mutate(split_set = case_when(
      between(s, 0.0, 0.6) ~ "train",
      between(s, 0.6, 0.8) ~ "test",
      between(s, 0.8, 1.0) ~ "validate"
    )) %>%
    select(-s) %>%
    mutate(across(c(gene, pert_iname_compound), as.factor)) %>%
    select(split_set, is_connected, gene, pert_iname_compound) %>%
    arrange(split_set, is_connected, gene, pert_iname_compound)
}
```


```{r}
get_training_bias_counts <- function(splits_i) {
  splits_i <-
    splits_i %>%
    filter(is_connected)

  gene_train_count <-
    splits_i %>%
    filter(split_set == "train") %>%
    count(gene, name = "n_connected_compounds_train", .drop = FALSE)

  compound_train_count <-
    splits_i %>%
    filter(split_set == "train") %>%
    count(pert_iname_compound, name = "n_connected_genes_train", .drop = FALSE)

  training_bias <-
    splits_i %>%
    filter(split_set == "test") %>%
    inner_join(gene_train_count, by = "gene") %>%
    inner_join(compound_train_count, by = "pert_iname_compound") %>%
    rowwise() %>%
    mutate(which_connected_train = case_when(
      (n_connected_compounds_train == 0) &
        (n_connected_genes_train == 0) ~ "neither",
      (n_connected_compounds_train == 0) &
        (n_connected_genes_train  > 0) ~ "gene",
      (n_connected_compounds_train >  0) &
        (n_connected_genes_train == 0) ~ "compound",
      (n_connected_compounds_train >  0) &
        (n_connected_genes_train  > 0) ~ "both"
    )) %>%
    ungroup() %>%
    mutate(which_connected_train = factor(which_connected_train, levels = c("both", "compound", "gene", "neither")))

  training_bias %>%
    count(which_connected_train, .drop = FALSE) %>% pivot_wider(names_from = "which_connected_train", values_from = n, values_fill = 0)

}
```

```{r}
training_bias_counts_genemod

splits_genemod %>%
  get_training_bias_counts()
```

```{r}
x0 %>%
  create_splits() %>%
  get_training_bias_counts()
```

```{r eval=TRUE}
training_bias_counts <-
  seq(1000) %>%
  map_df(~get_training_bias_counts(create_splits(x0)))

training_bias_counts %>%
  write_csv("output/training_bias_counts_simulations.csv")
```


```{r}
training_bias_counts <-
  read_csv("output/training_bias_counts_simulations.csv")
```


```{r}
training_bias_counts <-
  training_bias_counts %>%
  mutate(both_frac     = both     / (both + gene + compound + neither),
         compound_frac = compound / (both + gene + compound + neither),
         gene_frac     = gene     / (both + gene + compound + neither),
         neither_frac  = neither  / (both + gene + compound + neither))

training_bias_counts_genemod <-
  training_bias_counts_genemod %>%
  mutate(both_frac     = both     / (both + gene + compound + neither),
         compound_frac = compound / (both + gene + compound + neither),
         gene_frac     = gene     / (both + gene + compound + neither),
         neither_frac  = neither  / (both + gene + compound + neither))
```


```{r}
training_bias_counts %>%
  select(matches("frac")) %>%
  pivot_longer(everything(), names_to = "which_connected", values_to = "frac") %>%
  ggplot(aes(frac)) +
  geom_histogram(bins = 100) +
  geom_vline(data =
               training_bias_counts_genemod %>%
               select(matches("frac")) %>%
               pivot_longer(everything(),
                            names_to = "which_connected",
                            values_to = "frac"),
             aes(xintercept = frac),
             color = "red"
             ) +
  facet_wrap(~which_connected)
```

The counts we have in genemod are not rare after all (although `compound` is on the lower side, and `neither` on the higher side though)